---
title: "20240828_ML_Class_Exercises"
output: pdf_document
date: "2024-08-28"
---

1) The CARET package provides a built cell segmentation data set called `segmentationData`. To initialize access to the data, you can simply load the CARET package and call data(segmentationData). The loaded data frame will contain 119 imaging measurements on 2019 individual cells taken from a high content screening assay. The team that generated this data set sought to determine if the cells in the assay were segmented well or poorly. 

Read in this data and report how many outcome variables exist in the "Class" variable. 
```{r setup}

```
2) Now that you have determined how many outcomes are possible, split your data into 2/3 training data and 1/3 testing data. Make sure you conduct preprocessing before splitting the data! Note: the authors of the publication associated with this data provided a column called "Case" to this data frame indicating they used 50% of their data for training and the other 50% for testing. Remove this column before you conduct downstream processing. 
```{r}

```
3) Build an elastic net model that leverages recursive feature elimination to select a subset of important features to be included in your model. Use bootstrapping for cross validation. 
```{r}

```
4) Use your new model to predict whether a cell is poorly or well segmented in your testing data set. 
```{r}

```

