Good Day QBS 103 Crew!

Please work in groups to solve the problems provided in our very first in class assignment. If you complete this early, you are free to leave before our class time runs out. If you are unable to complete these exercises before the end of class, you can continue at home, though it is not required that you do so. These will not be graded!

1) Below I have generated a beautiful vector of "things" for you. Using code, can you tell me the type of data of each "thing" in the vector?
```{r setup, include=TRUE}
my_vector <- c(3.14, "pie", 3)
```

2) Create a nested vector in the code chunk below. This vector should contain one vector of integers, one vector of float numbers, and one vector of strings or characters.
```{r}

```

3) I have provided you with a vector of float numbers. Please convert these float numbers to integers and save the output vector to a vector called "my_integers".
```{r}
my_floats <- c(1.67, 1.11, 2.25, 8.88, 6.67, 104874587645776.2)
```

4)

```{r}
wilson_story <- c("Professor", "Kosarek's", "rabbit,","Wilson," "stumbled", "upon", "a","magnificent" "field" of "carrots." Excitement filled Wilson's "heart" as he hurried to share the news with his "rabbit" "friends." They gathered, their "noses twitching" with anticipation. Wilson, the "carrot-loving" host, declared a "carrot party." Laughter filled the air as the rabbits feasted on the "juicy" "delights." Crunching sounds echoed as they savored each "bite." Bellies full and spirits high, they bid their farewells. Wilson watched them "hop" away, feeling grateful for the "carrot-filled" memories. The field of carrots became a cherished "meeting" place, where joy and friendship bloomed with every "carrot party" they celebrated.)
```
