---
title: "In Class Solutions: Data Visualization"
author: "QBS 103"
date: "July 13, 2023"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### In Class Activity

Using the same data we've generated in class thus far, recreate the following plot:

``` {r message = F, warning = F}
# Load packages
library(tidyverse) # note: tidyverse includes ggplot2

# Set a random seed
set.seed(103)

# Define a data frame with our randomly generated data
randomData <- data.frame('SubjectID' = seq(1:1000), 
                         'Systolic.BP' = rnorm(n = 1000,mean = 128,sd = 20),
                         'Diastolic.BP' = rnorm(n = 1000,mean = 71,sd = 10),
                         'Age' = trunc(runif(n = 1000,min = 18,max = 70)),
                         'Male' = rbinom(n = 1000,size = 1,prob = 0.5))
# Define binary variable for biological sex
randomData$BiologicalSex <- factor(ifelse(randomData$Male == 1,'Male','Female'))
# Define variable specifying age above 65 (medicare eligible)
randomData$MedicareAge <- ifelse(randomData$Age < 65,F,T)
# Define a variable for hypertension
randomData$Hypertension <- ifelse(randomData$Systolic.BP > 130 | randomData$Diastolic.BP > 80,
                                  'Hypertensive','Normotensive')

# Make more categories in our data
randomData$AgeGroup <- cut(randomData$Age,breaks = c(0,30,40,50,60,70),
                           labels = c('Under 30','30-40','40-50','50-60','60-70'))
```

``` {r}

# Define theme
newTheme <- theme(
  # Remove all borders and grid lines
  panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank(), 
  # Set axis colors and size 
  axis.line = element_line(colour = "white", linewidth = rel(2)),
  # Change all colors to light blue
  plot.background = element_rect(fill = "black"),
  panel.background = element_blank(),
  legend.key = element_rect(fill = 'black'), 
  legend.background = element_rect(fill = 'black'),
  # Set font
  text = element_text(face="bold",colour = "white"),
  axis.text = element_text(face="bold",colour = "white"),
        legend.position = 'bottom')

# Define a color palette
colorPalette <- rev(wesanderson::wes_palette(c('Royal2'),n = 5))

# Calculate mean systolic bp
meanBP <- round(mean(randomData$Systolic.BP),digits = 2)

# Generate plot
ggplot(randomData,aes(x = Age,y = Systolic.BP,color = AgeGroup)) +
  geom_point() +
  scale_color_manual(values = colorPalette) +
  labs(x = "Age (yrs)",y = "Systolic Blood Pressure (mmHg)",color = "Age") +
  geom_vline(xintercept = 30, linetype = 'dashed',color = colorPalette[1]) +
  geom_vline(xintercept = 40, linetype = 'dashed',color = colorPalette[2]) +
  geom_vline(xintercept = 50, linetype = 'dashed',color = colorPalette[3]) +
  geom_vline(xintercept = 60, linetype = 'dashed',color = colorPalette[4]) +
  geom_vline(xintercept = 70, linetype = 'dashed',color = colorPalette[5]) +
  annotate(geom = 'text',x = 45,y = 220,label = paste('Mean Systolic Blood Pressure:',meanBP,'mmHg'),color = 'white') +
  newTheme

```

Hints: 

1. The color palette is "Royal2" from the *wesanderson* package but can you think of how you might still reproduce this without knowing that? **You can also find colors using different hex identifier resources.**

2. You can change the font in a theme using variations of "text" which you can find in the documentation of the *theme()* function.

3. Look up the documentation for *geom_hline* and *geom_vline*.