---
title: 'Math for Data Analytics I'
author: "QBS 103: Foundations of Data Science"
date: "August 20, 2024"
output: pdf_document
---
  
```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
```

### Lesson Objectives

##### At the end of this lecture you should be able to:

1. Read and interpret mathematical notation
2. Write and plot mathematical functions in R
3. Write logarithmic and trigonometric functions in R


### Resources

LaTeX Cheat Sheet for Set Notation: https://www.overleaf.com/learn/latex/List_of_Greek_letters_and_math_symbols 

Good Explanation of Mathematical Notation: https://www.youtube.com/watch?v=kaJuXx6uYR0 

LaTeX in R Plots Using *latex2exp*: https://cran.r-project.org/web/packages/latex2exp/vignettes/using-latex2exp.html  


### Mathematical Notation

Mathematical equations and proofs often rely on set notation to efficiently state information. This is a quick overview (not a comprehensive list) of some mathematical notation you might encounter as you move forward in statistics. 

For example, we can define $$A = \{x,y,z\}$$ which means that $A$ is a set that contains elements $x$, $y$, and $z$. We can refer to elements of the set using $\in$, such that we can say $$x \in A$$ 

Similarly, we can use the element symbol dashed through like this $\notin$ to indicate something isn't an element of that set, such that $$s \notin A$$

One special set that you might see is $\mathbb{R}$ which indicates the set of all real numbers.

We can also define a set as $$\{x|x > 0\}$$ rather than listing out individual values which would indicate the set of x in which x is greater than 0. 

We also use $\forall$ to indicate "for all" such that we can say $$ \forall \ x \in A, x > 0 $$
Reading across this just says, "for all x elements of A, x is greater than zero" or, more simply, all the elements of A are greater than zero.

We can also use the $\exists$ symbol to indicate "exists" such that we can say  $\forall \in A \ \exists$ such that $x > 0$
Reading across this means that there exists an element in A that is greater than zero. Similarly to $\notin$ we also have $\nexists$ which indicates "there does not exist".


### Simple Mathematical Functions in R

In this course already, we have been using functions, simulating data, making plots, and even using some linear algebra - all of which are foundational to calculus!
  
For instance, supposed we are interested in the following mathematical function: $$ f(x) = mx+b $$
  
To create this function in R, we represent this function as:
  
```{r}


```

And then if we were to solve this function for $m=1$, $x=2$, and $b=3$ we could: $$f(2)=1(2)+3=5$$ or, use our function:

```{r}


```

It is then also super easy to evaluate this function over a range of x values. For instance, we may want to know the solution of $f(x)$ over the domain $\{x \in  \mathbb{N} | 0\leq x \leq1000 \}$. You could solve this by hand, or you could generate values of x over your domain and solve like so:

```{r}


```

And you can see that we then have each solution for every value of x! Now we can easily plot these values.

```{r}


```

Representing functions this way becomes increasingly useful with increasingly complex functions. For instance, suppose we have the following function: $$f(x)=x\text{sin}\left(\frac{1}{x}\right)$$ and we want to solve it over the domain $\{x\in \mathbb{R} | -1\leq x \leq 1\}$. You could solve this by hand over many possibilities of x, or:

```{r}


```

Sometimes, you might want to use LaTeX within your plots. To do so, we'll use the *latex2exp* library. Additional information and examples are linked at the start of the lecture under resources.

```{r}

# Load package
#install.packages('latex2exp')
library(latex2exp)

# Increase margin size for function
# Syntax here is: par(mar = c(bottom, left, top, right) 
par(mar = c(5, 5, 4, 2) + 0.1) 



```

### Exponents amd Logarithms in R

In R, *log(x)* calculates the natural log of x (i.e. base-$e$). If you type in ?log you'll see that you have the ability to customize the base within the log function as well. 

```{r}



```

Given that $log_2$ and $log_10$ are both commonly used logarithms, there are separate functions for these as well.

```{r}


```

Given that $log_2$ and $log_10$ are both commonly used logarithms, there are separate functions for these as well. We've already gotten very used to using the ^ symbol for creating exponents. When we want to raise something to the power of $e$ we use the *exp()* function in R.


```{r}


```

You'll recall that there are special rules for using math with logarithms. You'll want to remember these before you dive back into statistics and calculus. We can also confirm these in R pretty easily. They include:

$$ log_{10}(ab) = log_{10}(a)+log_{10}(b)$$

```{r}



```

$$ ln(\frac{a}{b}) = ln(a)-ln(b)$$

```{r}

```

$$ln(1/x)=-ln(x)$$

```{r}


```

$$ log_2(x^y) = ylog_2(x)$$

```{r}

```

$$ln(e)=1$$

```{r}

```

$$ln(1)=0$$

```{r}

```

Similarly, there are a variety of exponent laws worth reviewing:

$$e^ae^b=e^{a+b}$$

```{r}

```

$$e^{ln(xy)}=xy$$

```{r}

```

$$2^{log_2(xy)}=xy$$

```{r}

```

$$\frac{e^a}{e^b}=e^{a-b}$$

```{r}

```

$$(e^a)^b=e^{ab}$$

```{r}

```

$$(xy)^a=x^ay^a$$

```{r}

```

$$x^{-a}=\frac{1}{x^a}$$

```{r}

```

$$x^{\frac{a}{b}}=\sqrt[b]{x^a}$$

```{r}

```


### Trigonometry

You'll often need to use trigonometry in Statistics and Calculus, so naturally our trig functions are built into R as well. Above, we used the *sin()* function in a plot but there are also functions for *cos()* and *tan()* as well as arcsin, arcos, and arctan, all shown below. 

```{r}

```

You'll reall that there are also a variety of trig identities you can use to make solving functions with trigonometry easier! A complete cheat sheet can be found here: <https://sciencenotes.org/trig-identities-study-sheet/> but below are some useful ones you'll frequently use:

$$ \tan\theta = \frac{\sin\theta}{\cos\theta}$$

```{r}

```

$$ \sin^2\theta+\cos^2\theta=1$$

```{r}

```

$$ \sin(2\theta)=2\sin\theta\cos\theta$$

```{r}

```

$$ \cos(2\theta)=\cos^2\theta-sin^2\theta = 2\cos^2\theta-1$$

```{r}

```

$$sin(-\theta)=-\sin(\theta)$$

```{r}

```

You may be wondering why we are reviewing so much of this in the context of R, and one very great reason is because R will be a useful tool in checking your work when you are required to use these laws to manually simplify functions. Consider for instance:

$$ f(\theta) = \frac{1+\sin\theta}{\cos\theta} - \frac{\cos\gamma}{1-\sin\gamma} $$

And you simplify to:

$$ f(\theta) = \frac{2\sin\theta-2\sin\gamma}{\sin(\theta-\gamma)+\cos\theta-cos\gamma} $$

We can test this in R:

```{r}

```